Left-linear rules in adhesive categories

### 2024-02-14 icalp/fusione.tex
pass generale, etc etc

\item Category of abstract derivation for DPO rewriting system $(\X, \R)$
  \begin{itemize}
  \item \emph{DPO-derivation category} $\dpo$: objects are graphs,
    arrows between possibly empty, derivations.
  \item \emph{DPO-abstract decorated derivation category} $\dpi$
  \end{itemize}

@@

\item Three-steps lemma (Lemma~\ref{lem:indep-global-left}), which is a weakened version of ``indepedence is global''.
  \begin{itemize}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph serves to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered and some new
entities can be generated. Concerning concurrency, the chosen notion
is shift equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Section~4]{Ehrig78}\cite[Chapter 3, Sec. 3.4]{CorradiniMREHL97}.

@@

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class of left-linear ones.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for future work.

\section{$\mathcal{M}$-adhesive categories and DPO rewriting systems}\label{sec:ade}
This first section recalls definitions and basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.

@@

  A \emph{left-linear DPO rewriting system} is a pair $(\X, \R)$ where
  $\X$ is an $\mathcal{M}$-adhesive category and $\R$ is a set of left
  $\mathcal{M}$-linear rules. The rewriting system $(\X, \R)$ is
  called \emph{linear}
  %
  \full{\emph{(consuming)}}
  %
  if every rule in $\R$ is
  $\mathcal{M}$-linear.
  \full{(consuming)}

@@

  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    The \emph{DPO-derivation graph} of
    $(\X, \R)$ is the (large) directed graph $\gpo$ having as vertices
    the objects of $\X$ and in which an edge between $G$ and $H$ is a
    direct derivation $\dder{D}\colon G\Mapsto H$. A \emph{derivation}
    $\der{D}$ between two objects $G$ and $H$ is a path between them
    in $\gpo$. The \emph{source} and \emph{target} of $\der{D}$ are,
    respectively, $G$ and $H$.
  \end{definition}
}

@@

\begin{definition}[Derivation]
  Let  $\X$ be an $\mathcal{M}$-adhesive category and 
  $(\X, \R)$ a left-linear DPO rewriting system. 
  Given $G, H \in \X$, a \emph{derivation}
  $\der{D}$ with source $G$ and target $H$, written
  $\der{D}: G \Mapsto H$, is defined as a sequence
  $\{\dder{D}_i\}_{i=0}^n$ of direct derivations such that
  \begin{enumerate}
  \item $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for
    every index $i$;
  \item $G_0=G$ and $G_{n+1}=H$.
  \end{enumerate}
  Moreover, we have an \emph{empty derivation} $G : G \Mapsto G$ for each $G \in \X$.
  %
  We will denote by $\lgh(\der{D})$ the \emph{length} of derivation $\der{D}$.
 %
  \full{We will call the number $n+1$ the \emph{length} of the derivation,
  denoted by $\lgh(\der{D})$. We will also say that an empty
  derivation has length $0$.}

@@

\noindent
\parbox{8cm}{
  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear DPO rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \{G_i,D_i\}}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, and use the symbol $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R_i \ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}

@@

Given DPO rewriting system $(\X, \R)$,  a derivation $\der{D}$ determines a diagram
$\Delta(\der{D})$ in $\X$. Using $\mathcal{M}$-adhesivity, it can be shown  that such diagram has a colimit $(\tpro{D}, \{\iota_X\}_{X\in \Delta(\der{D})})$ and that, if $(\X, \R)$ is linear, then every coprojection is in $\mathcal{M}$ ( see \Cref{lem:colim} for details and a proof of these statements).

@@

\begin{definition}[Consistent permutation]
  \label{def:permcon}
  Let $\X$ be an $\mathcal{M}$-adhesive category and consider a
  left-linear DPO rewriting system $(\X, \R)$ on it.  Take two
  non-empty decorated derivations $(\der{D}, \alpha, \omega)$ and
  $(\der{D}', \alpha', \omega')$ with the same length and with
  isomorphic sources and targets.

@@

We are now ready to prove the central result of this section.
\begin{lemma}\label{lem:impo}
  Let $(\X,R)$ be a left-linear DPO rewriting system and let also
  $(\der{D}_1, \alpha_1, \omega_1)$, $(\der{D}_2, \alpha_2, \omega_2)$
  be two composable decorated derivations of length, respectively,
  $l_1$ and $l_2$. Suppose that
  $\sigma:[0, l_1+l_2-1]\to [0, l_1+l_2-1]$ is a consistent
  permutation between
  $(\der{D}_1, \alpha_1, \omega_1)\cdot (\der{D}_2, \alpha_2,
  \omega_2)$ and
  $(\der{D}'_1, \alpha'_1, \omega'_1)\cdot (\der{D}'_2, \alpha'_2,
  \omega'_2)$ for some other two composable decorated derivations
  $(\der{D}_1, \alpha_1, \omega_1)$ and
  $(\der{D}_2, \alpha_2, \omega_2)$.  If
  $\tau:[0,l_1-1]\to [0, l_1-1]$ is a consistent permutation between
  $(\der{D}_1, \alpha_1, \omega_1)$ and
  $(\der{D}'_1, \alpha'_1, \omega'_1)$. Suppose that the set
  \[D_{\sigma, \tau}:=\{i\in [0, l_1-1]\mid \sigma(i)\neq \tau(i)\}\]
  is non-empty and let $j$ be its minimum. Let also $r(\der{D}_1)$ be $\{\rho_i\}_{i=0}^{l_1-1}$. Then the following hold true:
  \begin{enumerate}
  \item if $j=0$, then the rule $\rho_0$ is not consuming;
  \item if $j\neq 0$ then the rule $\rho_{j-1}$ is not consuming.
  \end{enumerate}
\end{lemma}
\begin{remark}\label{rem:minmax}It is worth to notice that the hypothesis $D_{\sigma, \tau} \neq \emptyset$ entails $l_1\neq 0$. Moreover, if $j\neq 0$, then $j-1$ is the maximum of $E_{\sigma, \tau}$.
\end{remark}

@@

\begin{corollary}[Uniqueness of consistent
  permutation]\label{cor:unique}
  Let $(\X,\R)$ be a consuming left-linear DPO rewriting system. Then,
  for every two decorated derivations $(\der{D}, \alpha, \omega)$ and
  $(\der{D}', \alpha', \omega')$, there exists at most one consistent
  permutation between them.
\end{corollary}
\begin{proof}
  Let
  $\sigma, \tau: [0, \lgh(\der{D})-1]\rightrightarrows [0,
  \lgh(\der{D})-1]$ be two consistent permutations. Let $H$ and $H'$
  be, respectively, the target of $\der{D}$ and $\der{D}'$. Take two
  isomorphisms $\gamma:\pi(H)\to \gamma$, $\gamma':\pi(H')\to H'$,
  then, according to \Cref{def:conc}:
  \begin{align*}(\der{D}, \alpha, \omega) & =(\der{D}, \alpha,
    \omega)\cdot (\der{D}_2, \gamma, \gamma) \\(\der{D}', \alpha',
    \omega')&=(\der{D}', \alpha', \omega')\cdot (\der{D}'_2, \gamma',
    \gamma')
  \end{align*}
  where $\der{D}_2$ and $\der{D}'_2$ are the empty derivation on $H$
  and $H'$. Since $(\X, \R)$ is consuming, then \Cref{lem:impo}
  entails $D_{\sigma, \tau}=\emptyset$, from which the thesis follows.
\end{proof}

\begin{lemma}\label{lem:presuffix} Let $(\X, \R)$ be a consuming
  left-linear DPO rewriting system.  Let also
  $(\der{D}_1, \alpha_1, \omega_1)$, $(\der{D}_2, \alpha_2, \omega_2)$
  be two composable decorated derivations of length, respectively,
  $l_1$ and $l_2$. Suppose that
  $\sigma\colon [0, l_1+l_2-1]\to [0, l_1+l_2-1]$ is a consistent
  permutation between
  $(\der{D}_1, \alpha_1, \omega_1)\cdot (\der{D}_2, \alpha_2,
  \omega_2)$ and
  $(\der{D}'_1, \alpha'_1, \omega'_1)\cdot (\der{D}'_2, \alpha'_2,
  \omega'_2)$ for some other two composable decorated derivations
  $(\der{D}_1, \alpha_1, \omega_1)$ and
  $(\der{D}_2, \alpha_2, \omega_2)$.  If
  $\tau\colon [0,l_1-1]\to [0, l_1-1]$ is a consistent permutation
  between $(\der{D_1}, \alpha_1, \omega_1)$ and
  $(\der{D}'_1, \alpha'_1, \omega'_1)$. Then the following hold true:
  \begin{enumerate}
  \item suppose that $l_1$ and $l_2$ are different from $0$ and
    consider the following two pushout squares given by
    \Cref{rem:dett}:
    \[\xymatrix@C=45pt@R=30pt{\pi(G_{2,0}) \ar@{>->}[d]_{\iota_{1,
            G_{1,n+1}} \circ \omega_1}\ar[r]^{\iota_{2, G_{2,0}} \circ
          \alpha_2}& \lpro \der{D}_2 \rpro \ar@{>->}[d]^{q_2} &
        \pi(G'_{2,0}) \ar@{>->}[d]_{\iota'_{1, G'_{1,n+1}} \circ
          \omega'_1} \ar[r]^{\iota'_{2, G'_{2,0}} \circ \alpha'_2}&
        \lpro \der{D}'_2 \rpro \ar@{>->}[d]^{q'_2} \\ \lpro \der{D}_1
        \rpro \ar[r]_{q_1} & \tpro{D} & \lpro \der{D}'_1\rpro
        \ar[r]_{q'_1} & \lpro \der{D}'\rpro }\] then, for every
    $i\in [0, l_2]$ there exists a unique
    $\zeta_i\colon G_{2,i}\to \lpro \der{D}_2\rpro $ fitting in the
    diagram below
    \[\xymatrix@R=35pt{G_{2,i} \ar[r]^{\iota_{2,G_{2,i}}} \ar@{.>}[d]_{\zeta_i}&\lpro \der{D}_2 \rpro \ar@{>->}[r]^{q_2}& \tpro{D} \ar[d]^{\xi_{\sigma}} \\
        \lpro \der{D}'_2 \rpro \ar@{>->}[rr]_{q'_2}&& \lpro
        \der{D}'\rpro }\]
  \item the permutation
    \[\varrho\colon [0,l_2-1]\to [0, l_2-1] \qquad i \mapsto
      \sigma(i+l_1)-l_1\] is a consistent one.
  \end{enumerate}
\end{lemma}

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  An \emph{independence
    pair} between $\dder{D}_0$ and $\dder{D}_1$, is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes
  \[\xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}\] In this case we say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  % \emph{weakly sequentially independent} if an independence pair
  % exists. If such independence pair is unique we will say that
  % $\dder{D}$ and $\dder{D'}$ are
  \emph{sequentially independent} and write $\dder{D}_0 \seqind \dder{D}_1$.
\end{definition}

@@

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear DPO rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ \todo{A: ma manca ``and  $\dder{D}_j = \dder{E}_j$ for each $j \not \in \{i,i+1\}$''??} we write
  $\der{D} \shift{\transp{i}} \der{E}$.

@@

\begin{definition}[strong independence pair]
  \label{def:filler}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear DPO rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=18pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@{.>}@/^.20cm/[ddrr]|(.32)\hole|(.57)\hole_(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole^(.2){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]_(.2){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@{.>}@/_.20cm/[ddll]|(.32)\hole|(.57)\hole^(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_(.3){g_0}&& G_1 &&
      \ar@{>->}[ll]^(.3){f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>.>}[ull]^{p_0} \ar@/_.3cm/@{.>}[urr]_{p_1}
    }
  \]

@@

\begin{definition}[Strong Enforcing rewriting systems]
  A left-linear DPO rewriting system is \emph{strong enforcing} if
  every independence pair between two direct derivations is strong.
\end{definition}

@@

Another interesting class of well-switching rewriting systems is given by the so-called E-graphs.

>>>>

\item Category of abstract derivation for rewriting system $(\X, \R)$
  \begin{itemize}
  \item \emph{DPO-derivation category} $\dpo$: objects are graphs,
    arrows between possibly empty, derivations.
  \item \emph{DPO-abstract decorated derivation category} $\dpi$
  \end{itemize}

@@

\item Three-steps lemma (Lemma~\ref{lem:indep-global-left}), which is a weakened version of ``independence is global''.
  \begin{itemize}

@@

Among rewriting-based formalisms, those manipulating graph-like
structures proved useful in many settings for representing the
dynamics of distributed systems: the graph serves to represent the
entities and their relations within states, while rewriting steps,
modifying the graph, model computation steps that result in changes to
the system state. The DPO (double-pushout) approach~\cite{EhrigPS73}
is nowadays considered a standard one for these structures, thanks to
its locality (rule application has a local effect on a state,
differently from more recent approaches like
SqPO~\cite{CorradiniHHK06} or PBPO$^+$~\cite{OverbeekER23}) and to its
flexibility: the rules allow to specify that, in a rewriting step,
some entities in the current state are removed, some others are needed
for the rewriting step to occur but remain unaltered and some new
entities can be generated. Concerning concurrency, the chosen notion
is shift equivalence, and independence is formally captured by what is
called the Church-Rosser theorem of DPO
rewriting~\cite[Chapter 3, Section~3.4]{CorradiniMREHL97}. %cite[Section~4]{Ehrig78}

@@

The paper has the following structure. 
Section~\ref{sec:ade} recalls the basic 
definitions of DPO rewriting for $\mathcal{M}$-adhesive categories. 
Section~\ref{sec:equi} presents the key novelties of our proposal,
first introducing the classical notion of independence and 
a novel axiomatic notion of switchability, further showing that the
former fails to imply the latter for left-linear rewriting systems.
We then define well-switching systems,
and prove that all the classical results holding for linear rewriting systems
can be recovered for such a class of left-linear ones.
Finally, Section~\ref{conclusions} wraps up the paper and 
offers some thoughts on possible directions for future work.
The appendixes collect additional data, recalling basic results on
adhesive categories (Appendix~\ref{app:ade}), introducing and 
proving properties for \emph{strong enforcing} rewriting systems
(Appendix~\ref{app:fill}), and giving the proofs of the main 
results of our work (Appendix~\ref{omitted}).

\section{$\mathcal{M}$-adhesive categories and rewriting systems}\label{sec:ade}
This first section recalls definitions and basic theory of \emph{$\mathcal{M}$-adhesive categories} \cite{azzi2019essence,ehrig2012,ehrig2014adhesive,lack2005adhesive,heindel2009category}.

@@

  A \emph{left-linear rewriting system} is a pair $(\X, \R)$ where
  $\X$ is an $\mathcal{M}$-adhesive category and $\R$ is a set of left
  $\mathcal{M}$-linear rules. The rewriting system $(\X, \R)$ is
  called \emph{linear}
  %
  \full{\emph{(consuming)}}
  %
  if every rule in $\R$ is
  $\mathcal{M}$-linear.
  \full{(consuming)}

@@

  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear rewriting system. 
    The \emph{DPO-derivation graph} of
    $(\X, \R)$ is the (large) directed graph $\gpo$ having as vertices
    the objects of $\X$ and in which an edge between $G$ and $H$ is a
    direct derivation $\dder{D}\colon G\Mapsto H$. A \emph{derivation}
    $\der{D}$ between two objects $G$ and $H$ is a path between them
    in $\gpo$. The \emph{source} and \emph{target} of $\der{D}$ are,
    respectively, $G$ and $H$.
  \end{definition}
}

@@

\begin{definition}[Derivation]
  Let  $\X$ be an $\mathcal{M}$-adhesive category and 
  $(\X, \R)$ a left-linear rewriting system. 
  Given $G, H \in \X$, a \emph{derivation}
  $\der{D}$ with source $G$ and target $H$, written
  $\der{D}: G \Mapsto H$, is defined as a sequence
  $\{\dder{D}_i\}_{i=0}^n$ of direct derivations such that
  \begin{enumerate}
  \item $\dder{D}_i : G_i \Mapsto G_{i+1}$ is a direct derivation for
    every index $i$;
  \item $G_0=G$ and $G_{n+1}=H$.
  \end{enumerate}
  Moreover, we have an \emph{empty derivation} $G : G \Mapsto G$ for each $G \in \X$.
  %
  We will denote by $\lgh(\der{D})$ the \emph{length} of derivation $\der{D}$.
 %
  \full{We will call the number $n+1$ the \emph{length} of the derivation,
  denoted by $\lgh(\der{D})$. We will also say that an empty
  derivation has length $0$.}

@@

\noindent
\parbox{8cm}{
  \begin{definition}
    Let  $\X$ be an $\mathcal{M}$-adhesive category and 
    $(\X, \R)$ a left-linear rewriting system. 
    An \emph{abstraction equivalence} between two derivations $\der{D}$
    and $\der{D'}$ with the same length and
    $r(\der{D})=r(\der{D}')$, is a family of isomorphisms
    $\{\phi_X\}_{X\in \{G_i,D_i\}}$ such that the diagram on the right commutes
    for every $i\in [0, \lgh(\der{D})]$. We
    will say that $\der{D}$ are $\der{D}'$ are \emph{abstraction
      equivalent}, and use the symbol $\der{D}\equiv_a \der{D}'$, if there exists an abstraction equivalence between
    them.
  \end{definition}
}
\parbox{4cm}{\vspace{-1.35em}$\xymatrix@C=40pt{G'_i&D'_i \ar[r]^{g'_i}
    \ar@{>->}[l]_{f'_i}&G'_{i+1}\\ L_i \ar[u]^{m'_i} \ar[d]_{m_i}& K_i
    \ar[u]^{k'_i} \ar[d]_{k_i} \ar[r]^{r_i} \ar@{>->}[l]_{l_i}
    &R_i \ar[u]^{h'_i} \ar[d]_{h_i}\\G_i
    \ar@/_.45cm/[uu]_(.35){\phi_{G_i}}|\hole&D_i\ar@{>->}[l]^{f_i}\ar@/_.45cm/[uu]_(.35){\phi_{D_i}}|\hole
    \ar[r]_{g_i}&G_{i+1}\ar@/_.45cm/[uu]_{\phi_{G_{i+1}}}}$}

@@

Given a rewriting system $(\X, \R)$,  a derivation $\der{D}$ determines a diagram
$\Delta(\der{D})$ in $\X$. Using $\mathcal{M}$-adhesivity, it can be shown  that such diagram has a colimit $(\tpro{D}, \{\iota_X\}_{X\in \Delta(\der{D})})$ and that, if $(\X, \R)$ is linear, then every coprojection is in $\mathcal{M}$ ( see \Cref{lem:colim} for details and a proof of these statements).

@@

\begin{definition}[Consistent permutation]
  \label{def:permcon}
  Let $\X$ be an $\mathcal{M}$-adhesive category and consider a
  left-linear rewriting system $(\X, \R)$ on it.  Take two
  non-empty decorated derivations $(\der{D}, \alpha, \omega)$ and
  $(\der{D}', \alpha', \omega')$ with the same length and with
  isomorphic sources and targets.

@@

We are now ready to prove the central result of this section.
\begin{lemma}\label{lem:impo}
  Let $(\X,R)$ be a left-linear rewriting system and let also
  $(\der{D}_1, \alpha_1, \omega_1)$, $(\der{D}_2, \alpha_2, \omega_2)$
  be two composable decorated derivations of length, respectively,
  $l_1$ and $l_2$. Suppose that
  $\sigma:[0, l_1+l_2-1]\to [0, l_1+l_2-1]$ is a consistent
  permutation between
  $(\der{D}_1, \alpha_1, \omega_1)\cdot (\der{D}_2, \alpha_2,
  \omega_2)$ and
  $(\der{D}'_1, \alpha'_1, \omega'_1)\cdot (\der{D}'_2, \alpha'_2,
  \omega'_2)$ for some other two composable decorated derivations
  $(\der{D}_1, \alpha_1, \omega_1)$ and
  $(\der{D}_2, \alpha_2, \omega_2)$.  If
  $\tau:[0,l_1-1]\to [0, l_1-1]$ is a consistent permutation between
  $(\der{D}_1, \alpha_1, \omega_1)$ and
  $(\der{D}'_1, \alpha'_1, \omega'_1)$. Suppose that the set
  \[D_{\sigma, \tau}:=\{i\in [0, l_1-1]\mid \sigma(i)\neq \tau(i)\}\]
  is non-empty and let $j$ be its minimum. Let also $r(\der{D}_1)$ be $\{\rho_i\}_{i=0}^{l_1-1}$. Then the following hold true:
  \begin{enumerate}
  \item if $j=0$, then the rule $\rho_0$ is not consuming;
  \item if $j\neq 0$ then the rule $\rho_{j-1}$ is not consuming.
  \end{enumerate}
\end{lemma}
\begin{remark}\label{rem:minmax}It is worth to notice that the hypothesis $D_{\sigma, \tau} \neq \emptyset$ entails $l_1\neq 0$. Moreover, if $j\neq 0$, then $j-1$ is the maximum of $E_{\sigma, \tau}$.
\end{remark}

@@

\begin{corollary}[Uniqueness of consistent
  permutation]\label{cor:unique}
  Let $(\X,\R)$ be a consuming left-linear rewriting system. Then,
  for every two decorated derivations $(\der{D}, \alpha, \omega)$ and
  $(\der{D}', \alpha', \omega')$, there exists at most one consistent
  permutation between them.
\end{corollary}
\begin{proof}
  Let
  $\sigma, \tau: [0, \lgh(\der{D})-1]\rightrightarrows [0,
  \lgh(\der{D})-1]$ be two consistent permutations. Let $H$ and $H'$
  be, respectively, the target of $\der{D}$ and $\der{D}'$. Take two
  isomorphisms $\gamma:\pi(H)\to \gamma$, $\gamma':\pi(H')\to H'$,
  then, according to \Cref{def:conc}:
  \begin{align*}(\der{D}, \alpha, \omega) & =(\der{D}, \alpha,
    \omega)\cdot (\der{D}_2, \gamma, \gamma) \\(\der{D}', \alpha',
    \omega')&=(\der{D}', \alpha', \omega')\cdot (\der{D}'_2, \gamma',
    \gamma')
  \end{align*}
  where $\der{D}_2$ and $\der{D}'_2$ are the empty derivation on $H$
  and $H'$. Since $(\X, \R)$ is consuming, then \Cref{lem:impo}
  entails $D_{\sigma, \tau}=\emptyset$, from which the thesis follows.
\end{proof}

\begin{lemma}\label{lem:presuffix} Let $(\X, \R)$ be a consuming
  left-linear rewriting system.  Let also
  $(\der{D}_1, \alpha_1, \omega_1)$, $(\der{D}_2, \alpha_2, \omega_2)$
  be two composable decorated derivations of length, respectively,
  $l_1$ and $l_2$. Suppose that
  $\sigma\colon [0, l_1+l_2-1]\to [0, l_1+l_2-1]$ is a consistent
  permutation between
  $(\der{D}_1, \alpha_1, \omega_1)\cdot (\der{D}_2, \alpha_2,
  \omega_2)$ and
  $(\der{D}'_1, \alpha'_1, \omega'_1)\cdot (\der{D}'_2, \alpha'_2,
  \omega'_2)$ for some other two composable decorated derivations
  $(\der{D}_1, \alpha_1, \omega_1)$ and
  $(\der{D}_2, \alpha_2, \omega_2)$.  If
  $\tau\colon [0,l_1-1]\to [0, l_1-1]$ is a consistent permutation
  between $(\der{D_1}, \alpha_1, \omega_1)$ and
  $(\der{D}'_1, \alpha'_1, \omega'_1)$. Then the following hold true:
  \begin{enumerate}
  \item suppose that $l_1$ and $l_2$ are different from $0$ and
    consider the following two pushout squares given by
    \Cref{rem:dett}:
    \[\xymatrix@C=45pt@R=30pt{\pi(G_{2,0}) \ar@{>->}[d]_{\iota_{1,
            G_{1,n+1}} \circ \omega_1}\ar[r]^{\iota_{2, G_{2,0}} \circ
          \alpha_2}& \lpro \der{D}_2 \rpro \ar@{>->}[d]^{q_2} &
        \pi(G'_{2,0}) \ar@{>->}[d]_{\iota'_{1, G'_{1,n+1}} \circ
          \omega'_1} \ar[r]^{\iota'_{2, G'_{2,0}} \circ \alpha'_2}&
        \lpro \der{D}'_2 \rpro \ar@{>->}[d]^{q'_2} \\ \lpro \der{D}_1
        \rpro \ar[r]_{q_1} & \tpro{D} & \lpro \der{D}'_1\rpro
        \ar[r]_{q'_1} & \lpro \der{D}'\rpro }\] then, for every
    $i\in [0, l_2]$ there exists a unique
    $\zeta_i\colon G_{2,i}\to \lpro \der{D}_2\rpro $ fitting in the
    diagram below
    \[\xymatrix@R=35pt{G_{2,i} \ar[r]^{\iota_{2,G_{2,i}}} \ar@{.>}[d]_{\zeta_i}&\lpro \der{D}_2 \rpro \ar@{>->}[r]^{q_2}& \tpro{D} \ar[d]^{\xi_{\sigma}} \\
        \lpro \der{D}'_2 \rpro \ar@{>->}[rr]_{q'_2}&& \lpro
        \der{D}'\rpro }\]
  \item the permutation
    \[\varrho\colon [0,l_2-1]\to [0, l_2-1] \qquad i \mapsto
      \sigma(i+l_1)-l_1\] is a consistent one.
  \end{enumerate}
\end{lemma}

@@

\begin{definition}[Sequential independence]
  \label{de:sequential-independence}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also
  $\der{D}=\{\dder{D}_i\}_{i=0}^1$ be a derivation of length $2$, with
  $\dder{D}_i$ using rule $\rho_i = (l_i,r_i)$.  An \emph{independence
    pair} between $\dder{D}_0$ and $\dder{D}_1$, is a pair of arrows
  $i_0\colon R_0\to D_1$ and $i_1\colon L_1\to D_0$ such that the
  following diagram commutes
  \[\xymatrix@C=15pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} & R_0
      \ar@{.>}@/^.35cm/[drrr]|(.3)\hole_(.4){i_0}
      \ar[dr]|(.3)\hole_{h_0} && L_1 \ar@{.>}@/_.35cm/[dlll]^(.4){i_1}
      \ar[dl]|(.3)\hole^{m_1}& K_1 \ar[d]^{k_1}\ar@{>->}[l]_{l_1}
      \ar[rr]^{r_1} && R_1 \ar[d]^{h_1}\\G_0 && \ar@{>->}[ll]^{f_0}
      D_0 \ar[rr]_{g_0}&& G_1 && \ar@{>->}[ll]^{f_1} D_1
      \ar[rr]_{g_1}&& G_2}\] In this case we say that $\dder{D}_0$ and
  $\dder{D}_1$ are
  % \emph{weakly sequentially independent} if an independence pair
  % exists. If such independence pair is unique we will say that
  % $\dder{D}$ and $\dder{D'}$ are
  \emph{sequentially independent} and write $\dder{D}_0 \seqind \dder{D}_1$.
\end{definition}

@@

\begin{definition}[Switch equivalence]
  \label{de:switch-equivalence}
  Let $(\X, \R)$ be a left-linear rewriting system.  Let also
  $\der{D}, \der{E} : G \Mapsto H$ be derivations with the same
  length, $\der{D}=\{\dder{D}_{i}\}_{i=0}^n$ and
  $\der{E}=\{\dder{E}_{i}\}_{i=0}^n$. If
  $\dder{D}_i \cdot \dder{D}_{i+1}$ is a switch of
  $\dder{E}_i \cdot \dder{E}_{i+1}$ for some $i \in [0,n-1]$ \todo{A: ma manca ``and  $\dder{D}_j = \dder{E}_j$ for each $j \not \in \{i,i+1\}$''??} we write
  $\der{D} \shift{\transp{i}} \der{E}$.

@@

\begin{definition}[strong independence pair]
  \label{def:filler}
  Let $\X$ be an $\mathcal{M}$-adhesive category 
  and $(\X, \R)$ a left-linear rewriting system. 
  Let also $(i_0, i_1)$ be an independence pair between two direct
  derivations $\dder{D}_0$, $\dder{D}_1$
  as in the solid part of the diagram below
  \[
    \xymatrix@C=18pt{L_0 \ar[d]_{m_0}&& K_0
      \ar[d]_{k_0}\ar@{>->}[ll]_{l_0} \ar[r]^{r_0} \ar@{.>}@/^.20cm/[ddrr]|(.32)\hole|(.57)\hole_(.7){u_0} & R_0
      \ar@/^.35cm/[drrr]|(.3)\hole^(.2){i_0} \ar[dr]|(.3)\hole_{h_0}
      && L_1 \ar@/_.35cm/[dlll]_(.2){i_1} \ar[dl]|(.3)\hole^{m_1}& K_1
      \ar[d]^{k_1}\ar@{>->}[l]_{l_1} \ar[rr]^{r_1} \ar@{.>}@/_.20cm/[ddll]|(.32)\hole|(.57)\hole^(.7){u_1} && R_1 \ar[d]^{h_1} \\
      %
      G_0 &&
      \ar@{>->}[ll]^{f_0} D_0 \ar[rr]_(.3){g_0}&& G_1 &&
      \ar@{>->}[ll]^(.3){f_1} D_1 \ar[rr]_{g_1}&& G_2\\
      %
      & & & & P \ar@/^.3cm/@{>.>}[ull]^{p_0} \ar@/_.3cm/@{.>}[urr]_{p_1}
    }
  \]

@@

\begin{definition}[Strong Enforcing rewriting systems]
  A left-linear rewriting system is \emph{strong enforcing} if
  every independence pair between two direct derivations is strong.
\end{definition}

@@

Another interesting class of well-switching rewriting systems is given by e-graphs.
